Improving Emotion Detection through Translation of Text to ML Models Trained in Different Languages

### 2023-08-16 paper/main.tex
Paper Updates - Feedback Review

\begin{abstract}
This Qualifying Exam\footnote{Comp. and Data Science PhD students are required to complete a Qualifying Exam in their first year} research paper investigates the potential of improving Emotion Detection (ED) through translation of extended text data to multiple Machine Learning (ML) models trained in different languages. Our research focused on  English (X\footnote{X is formerly known as Twitter} threads) and German (voting articles) datasets to improve prediction accuracy. By this our research aims to address the challenges posed by the lack of comprehensive labeled datasets and language fragmentation in ED research and projects.

By extending an original English language dataset with a translated dataset from German to English the training data (extended dataset) becomes larger and the hope is that better prediction rates can be achieved in ED analysis applications due to ML training having more data for learning. Furthermore, translating English to German to extend the German dataset and training another ML model using PySpark and accessing both models in real-time by a simple RESTful API may improve the prediction rates even more by dual processing a sentence at the same time.

In order to present the results presented both orally to the MTSU's Computational Data \& Science Committee and within this paper labeled datasets both in English and German were collected, parsed, cleaned, translated, and two ML models built that can be accessed via an API to get a prediction by inputting sentences both in German or English via GET method using a simple query-string.

Sadly the research and results point to the realization that extending the datasets by translation did not improve prediction rates of English or German. Some minor gains can be achieved by accessing both English and German models at the same time via the RESTful API but unfortunately the benefits may not really support the efforts. In conclusion it that much more data needs to be collected in order to be able to definitely answer this research question.
\end{abstract}
\clearpage

@@

Due to the before mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions. Our research in turn contributes to the advancement of artificial intelligence, natural language processing (NLP) and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions} to go about our daily lives.

@@

\subsection{Challenges: Lack of Labeled Datasets and Language Fragmentation}
Emotion Detection is a fast growing field\cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}. The primary reason of the lack of data stems from the face that emotion detection data requires primarily supervised learning data. In in turn requires time and effort to construct, catalog, and procure,; this all is mostly created and labeled by humans. To compound on the issue of the lack of labeled data; the many datasets that are available are in many cases in multiple languages - not all are in English since emotions that are linked to text are contextual in nature. for this research a a significant amount of time took place to find the German labeled data, which eventually was found online by google searching.

@@

With the before mentioned issues regarding the quantity (lack of) and language fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED. The aim of the research is that by use of translation that the predictability of ED can be improved by extending an English dataset with translations from German and also process look-ups in real-time language conversion and processing.

@@

For instance, a study by Kapoor\cite{KAPOOR2023120882} et al. proposes analyzing the entire spectrum of the language source to predict emotional changes. This provided some help, but quickly established itself as time consuming task and need of expertise in the subject matters that could not outweigh the results. Overcoming the scarcity of labeled datasets and embracing innovative approaches like the before mentioned study by Kapoor\cite{KAPOOR2023120882} et al. and their analysis can contribute to the advancement of ED research. Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains, although the cost of such might be prohibitive at this time.

@@

It is important to note that in order to accurately test and compare the original and the extended datasets to each other we made sure that the test data was the same. to do this we randomly extracted 15\% of the original datasets (English and German) before extending the original with the translated datasets.

@@

\subsection{Findings and Implications for Improving Emotion Detection}
Based on the improvement results in Sections \ref{sec:english-dataset-results} and \ref{sec:german-dataset-results} it seems that using translation to extend both the English and German datasets resulted in a negative outcomes. The best outcomes were using a simple Linear Regression model on the English and German, however based on the F1 score (EQ \ref{eq:f1-score}) of the Random Forest Classifier it seems to perform the best on both English and German.

Unfortunately based on the results of extending the data our approach to also using real-time translation for multi-language models

\subsection{Conclusion and Future Work}
It's crucial to ensure the quality and accuracy of extended data, especially when language translation is involved, to prevent introducing more noise than meaningful information into the training process. Although the research did not garner any significant improvements by translating the German to English as a means on extending the English dataset, it actually made it the prediction results worse.

Future work includes testing and translating different models by using different translation services. Using Google Cloud's translation service was adequate for this research, however there are certainly better translation services that are available but are based on paid subscriptions.

>>>>

\begin{abstract}
This Qualifying Exam\footnote{Comp. and Data Science PhD students are required to complete a Qualifying Exam in their first year} research paper investigates enhancing Emotion Detection (ED) by translating extended text data into various Machine Learning (ML) models trained in distinct languages. We focused on English and German text data to enhance prediction accuracy, aiming to overcome challenges arising from limited labeled datasets and language fragmentation in ED research.

Expanding an original English dataset with translated German data increases the training data's volume, potentially improving prediction rates in ED applications. Additionally, translating English to German to extend the German dataset and accessing in real-time ML models trained in both could further improve prediction rates.

For presentation purposes, datasets in both English and German were collected, parsed, cleaned, and translated. Multiple ML models trained in English and German where built, and made accessible via an API for predictions in either language including real-time translation using a GET method.

The research findings suggest that the extension of datasets through translation has not yielded improvements in predictive accuracy for both English and German languages. Modest enhancements could potentially be achieved by concurrently accessing English and German models through real-time translation via the RESTful API; however, the benefits may not fully justify the efforts.

We posit that the prevalence of numerous classes in this multi-class classification model has contributed to instances of overfitting across several labels/classes. This occurrence, in turn, has led to a memorization effect rather than facilitating genuine learning within the models.

In conclusion, it becomes evident that a more substantial accumulation of data is required, or an innovative approach involving the utilization of AI to generate analogous data must be employed to comprehensively address this research question.
\end{abstract}
\clearpage

@@

Due to the above mentioned wide scope of uses for ED this field of research and the work to advance it provides significance and importance in enabling machines to comprehend and respond appropriately to human interactions. Our research in turn contributes to the advancement of artificial intelligence, natural language processing (NLP) and in many cases the utilization of text-to-speech applications most all of us in today's society use and rely on \cite{ai-framework-detection-emotions} to go about our daily lives.

@@

\subsection{Challenges: Lack of Labeled Datasets and Language Fragmentation}
Emotion Detection is a fast growing field\cite{ACLU-ED-Data, ACLU-THE-DAWN-OF-ROBOT-SURVEILLANCE}, however unlike Sentiment Analysis (SA) the availability of large datasets for training purposes of ML models is much smaller \cite{ACLU-ED-Data, ai-framework-detection-emotions}. The primary reason of the lack of data stems from the face that emotion detection data requires primarily supervised learning data. In turn requires time and effort to construct, catalog, and procure,; this all is mostly created and labeled by humans. To compound on the issue of the lack of labeled data; the many datasets that are available are in many cases in multiple languages - not all are in English since emotions that are linked to text are contextual in nature. for this research a a significant amount of time took place to find the German labeled data, which eventually was found online by google searching.

@@

With the above mentioned issues regarding the quantity (lack of) and language fragmentation of the ED datasets this research was motivated by finding a way to combine different language datasets into a single set for training processes to improve the predictability of ED. The aim of the research is that by use of translation that the predictability of ED can be improved by extending an English dataset with translations from German and also process look-ups in real-time language conversion and processing.

@@

For instance, a study by Kapoor\cite{KAPOOR2023120882} et al. proposes analyzing the entire spectrum of the language source to predict emotional changes. This provided some help, but quickly established itself as time consuming task and need of expertise in the subject matters that could not outweigh the results. Overcoming the scarcity of labeled datasets and embracing innovative approaches like the above mentioned study by Kapoor\cite{KAPOOR2023120882} et al. and their analysis can contribute to the advancement of ED research. Such advancements hold the potential to enhance the accuracy and applicability of emotion detection models, enabling a deeper understanding of human emotions and more effective responses in various domains, although the cost of such might be prohibitive at this time.

@@

It is important to note that in order to accurately test and compare the original and the extended datasets to each other we made sure that the test data was the same on each case. To do this we randomly extracted 15\% of the original datasets (English and German) and saved it as a test dataset prior to extending the original with the translated sentence data.

@@

\subsection{Findings and Implications for Improving Emotion Detection}
Based on the improvement results in Sections \ref{sec:english-dataset-results} and \ref{sec:german-dataset-results} it seems that using translation to extend both the English and German datasets resulted in a negative outcomes. The best outcomes were using a simple Linear Regression model on the English and German, however based on the F1 Score (Eq: \ref{eq:f1-score}) of the Random Forest Classifier it seems to perform the best on both English and German. This is because the false positive amount of one score on the accuracy is not taken into account and can easily skew the results when using a multi-class classification model.

Considering the outcomes of data extension via translation, our approach of utilizing real-time translation for multi-language models through the creation of an API and translating text in real-time for both German and English appears to be unsatisfactory in terms of achieving any substantial improvement as well.

With both these findings, we believe that the data of only using $\sim$1,500 (original dataset) and $\sim$2,700 (extended dataset) to train models is insufficient\cite{overfitting-overview} when using a multi-class classification problem. By inference, if we have only 1,500 and divide (eq: \ref{eq:sentence-per-class}) those by our eight (Table: \ref{table:dataset_labels}) classes we simply do not have enough data to truly train the ML on the classes.

\begin{equation}
\begin{split}
\frac{Features}{Labels}= \frac{Sentences}{Classes}=\frac{1,500}{8}=185 \\
or=\frac{2,700}{8}=337
\end{split}
\label{eq:sentence-per-class}
\end{equation}

Taking into account the insufficiency of data, the implication of a lack of classified data can lead to several significant issues for our ML models. Firstly, there's an absence of context for the models to operate effectively and grasp the intricate nuances of both the English and German languages. Secondly, in cases where class data is lacking, the models tend to memorize existing sentences and words, rather than learning the underlying patterns of the language. Both of these issues are connected to a phenomenon called overfitting\cite{overfitting-overview}, which is prevalent in many regression problems.

Options for improvement range from the straightforward approach of acquiring more data to the more nuanced strategy of leveraging existing labeled sentences and employing machines to generate derivatives. This would facilitate the creation of more intricate patterns for the machines to learn from.

\subsection{Conclusion and Future Work}
It's crucial to ensure the quality and accuracy of extended data, especially when language translation is involved, to prevent introducing more noise than meaningful information into the training process. The implications of lack of data in ED analysis presented it self in this research that it wanted to try and overcome. Although the research did not garner any significant improvements by translating the German to English as a means on extending the English dataset, it actually made it the prediction results worse. We believe one of the main culprits of this that the data for learning based on eight classes was insufficient. As can be seen in Equation \ref{eq:sentence-per-class} from the previous section, the models seemingly resorted to memorizing the sentence rather than learning the intricacies of the labeled classes. This inference is drawn that certain labels exhibited better F1 Score performance, primarily due to their greater availability in the training data.

Two areas of prospective research emerge: firstly, conducting further tests involving the translation of different models using diverse translation services. While Google Cloud's translation service sufficed for this research, superior paid subscription-based translation services are available. With the help of more data and better translation more concrete answers to prediction improvement may be achieved. Secondly, another avenue to overcome the lack of data involves employing AI to generate sentences that are similar yet distinct from the original. This could be achieved by providing a sentence and its associated labeled class to the AI for data creation. This approach aims to produce additional sentences imbued with similar emotional attributes, potentially expanding the labeled dataset.
